% -*- TeX:UK -*-
\documentclass[10pt]{beamer}
\usetheme{metropolis}
%\useinnertheme{rectangles}
\setbeamercovered{%
still covered={\opaqueness<1->{15}},
again covered={\opaqueness<1->{40}}}

\hypersetup{colorlinks,linkcolor=black,urlcolor=brown,citecolor=brown}

\include{shared-tex/set-up-fonts-and-icons}

\begin{document}

\title{PBIO-141\\Sensory and Physiological Ecology\\ of  Plants}
\subtitle{3: Signals, cues, information and evolution}
\author{Pedro J. Aphalo}
\date{January--February 2022}
\institute[Univ.\ of Helsinki]{M.Sc.\ in Plant Biology, University of Helsinki\\[2ex] \url{http://blogs.helsinki.fi/aphalo/}}

  \begin{frame}
    \maketitle
  \end{frame}

\include{shared-tex/copyright-frame}

  \begin{frame}
    \frametitle{Outline}
    \tableofcontents
  \end{frame}

\section{Experiments and surveys}

\begin{frame}{Types of experiments (refresher)}
\begin{small}
    \begin{description}[type=1]
        \item[Manipulative experiments.] With manipulative experiments it
        is possible to \emph{directly} demonstrate cause-effect relationships.

        \item[Observational experiments.] When manipulative experiments
        are not possible, direct demonstration of cause-effect
        relationships can be very difficult.

        \item[Impossible experiments.] In many situations
        manipulative experiments are physically or ethically
        impossible to carry out.

        \item[Simulations.] When we do experiments with models
        instead of the real system we call them simulations.

        \item[\DExamples]

    \end{description}
\end{small}
\end{frame}

\begin{frame}{Experiments and causality}
\begin{itemize}
\item Demonstration of causes. In manipulative experiments we (attempt to) keep
everything equal among treatments except for the experimental
variable whose effect we want to test.

\item In observational studies we cannot be sure of which things
are different among the different situations (``treatments''), so we
can make guesses about causes, but can seldom directly demonstrate them.

\item In observational studies we can use path analysis to indirectly strengthen the evidence for causation (we add an additional dimension, usually time).

\item \DExamples

\end{itemize}
\end{frame}

\begin{frame}{Cause vs.\ correlation}

Tight correlation does not demonstrate causality, two responses may
be dependent on a third variable and be correlated without one being
the cause of the other. \vspace{5mm} \pause

\emph{Example:} correlation between number of swimmers drowning in
lakes, and sales of ice cream. Both peak in the summer, but there is
no cause-effect relationship between them.

\emph{Possible way out:} path analysis. E.g.\ have dead swimmers eaten
ice cream just before drowning? (more strictly, \ldots more frequently than survivors).

\end{frame}

\begin{frame}{Cause and effect in Biology \Discussion 10 + 5 min}

\begin{centering}
$\to$ \emph{fitness} $\to$ genome $\to$  epigenome $\to$ phenome $\to$ \emph{fitness} $\to \cdots$
\vspace{3ex}
\end{centering}

\begin{description}
%  \item[time sequence] .
  \item[\DiscussionI] Which transitions, shown as $\to$, depend directly on the environment?
  \item[\DiscussionI] ``Evolution determines physiology'' or ``Physiology determines fitness''?
  \item[\Attention] The course of evolution is not the result of natural selection alone!
\end{description}
\pause
\Attention \emph{Multiple viewpoints are possible, and not necessarily contradictory!}
\end{frame}


  \section*{References}
  \begin{frame}[t,allowframebreaks]
    \frametitle{References}
    \printbibliography
  \end{frame}

\end{document}
